Psychophysics of musical elements in discrete-time representation of sound

### 2013-04-22 article/phi+mus.tex
artigo escrito ateh parte dos timbres

\section{Characterization of the discrete-time musical note}
In diverse artistic and theoretical contexts, music is though of as being constituted by units called notes and this units taken as "atoms" that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
Today, this notes are undestood as a central element of certain musical paradigms. In a cognitive sense, the notes are seen as discretizations that eases and enrich the flow of information through music~\cite{Roederer, Lacerda}
Canonically, a musical note posses at least duration, colume, pitch and tibre~\cite{Lacerda}. These are qualities which can be maneged quantitatively dictated by the evenly time spaced sound samples.

@@

\subsection{Duration}
The sample frequency $f_s$ is defined as the number os samples in each second of the discrete-time signal. Be $T_i=\{t_i\}$ an ordered set of real samples separated by $\delta_s=1/f_s$ seconds. A Musical note od duration $\Delta$ seconds is presented as a sequence of $\lfloor \Delta . f_s \rfloor $ samples. That is, taken the integer part of the multiplication, it is admited an error of at most $-\lambda_a$ seconds, which, for musical porpuses, are usually fine, for $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds. It is reasonable to state:

@@

\subsection{Volume}\label{subsec:volume}
The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics worked on section~\ref{sec:varInternas}. One can get variations on volume through the potency of the wave~\cite{Chowning}:

@@

\begin{enumerate}
  \item{Given a subject (or area) to be studied, objects belonging to
    this area are identified. Here, philosophers and composers were
    chosen arbitrarily for the fields of philosophy and classical
    music, respectively. In other applications, the choice could be
    objective; for example, in a study of metropolitan areas the
    choice of the objects (cities) would obey a very precise criterion
    (population).}

  \item{A set of attributes is established, which are used to
    characterize the objects. In our case, a few attributes were
    chosen arbitrarily based on well-known characteristics in the
    fields of philosophy and music. In order to make the analysis by
    humans feasible, the number of attributes had to be low. But in
    other types of work, the set could be chosen based on objective
    criteria and the number of attributes could be large.}

  \item{For a quantitative analysis, scores are assigned for each of
    the attributes to each of the objects. In this study, three of the
    authors assigned scores based on their knowledge of the fields
    under study. This can be generalized, with scores assigned with
    objective criteria. Moreover, schemes may be applied to check the
    quality of the assignment, for example using the Kappa
    index~\cite{Cohen1960} to verify agreement among the people who
    assigned the scores.}

  \item{The data generated from the steps above may be sparse and in
    small amount, therefore unsuitable for the application of
    statistical methods. In order to overcome this limitation, we
    introduced a step to verify the robustness of the analysis, which
    consists in generating ``artificial data'' via a bootstrap
    method~\cite{Boos2003}. It is worth noting that in other
    applications, this creation of artificial data may not be
    needed. For instance, we could have taken data for a much larger
    number of philosophers and composers. However, in this paper this
    would hamper the task of manual score assignment, and
    interpretation of the final results would be much more difficult
    (with so many names to be analyzed).}

  \item{Because the study is comparative, metrics must be found to
    identify similarities (or dissimilarities) among the objects, and
    projection methods are applied to visualize the data. There are
    several possibilities for this task, and here we used Pearson
    correlation and principal component analysis~\cite{Costa} to
    analyze the distances among objects. It served to establish a time
    line, with which the evolution of philosophy and music could be
    studied. Furthermore, with this time line it was possible to
    analyze the data in terms of concepts such as dialectics and
    opposition, which were obtained in a quantitative manner.}
\end{enumerate}

With regard to the specific methodology in the present work, we first
identified prominent music composers and philosophers along history,
and established a set of main musical and philosophical
features. Grades were then assigned to each of the composers and
philosophers for all features. The assignment of scores was not
arbitrary, for they were based on research about techniques and styles
used by composers and philosophers. The grades reveal a \emph{tendency
  index} on characteristics of composers or philosophers. For example,
to say that Bach is more contrapuntist than the other composers
corresponds to assigning a higher grade --- e.g.\ 8.0 or 9.0 --- to
the Barroque master and smaller grades to others. We chose a reduced
set of philosophers and composers for the sake of simplicity and
clarity. Then, to avoid statistical bias owing to the small number of
samples, we developed a bootstrap method~\cite{Varian} with which a
larger data set of 1000 new \emph{artificial composers and
  philosophers} were generated, being directly influenced by the
original sample and representing the contemporaries of the
philosophers and composers chosen.

The scores assigned to the characteristics of each composer --- and
philosopher --- define a state vector in its feature space. This
quantification of composers and philosophers allowed the application
of sound quantitative concepts and methods from multivariate
statistics~\cite{Papoulis, Wichern, Therrien} and pattern
recognition~\cite{Duda, Costa}. Correlations between these
characteristic vectors were identified and principal component
analysis (PCA)~\cite{Costa} was applied to represent the music --- and
philosophy --- history as a planar space where development may be
followed as vectorial movements. On this planar space, concepts like
dialectics, innovation and opposition, originally non-quantitative,
can be modeled as mathematical relations between individual states.

It is important to note that application of statistical analysis to
music is not recent. In musicology, statistical methods have been used
to identify many musical
characteristics. Simonton~\cite{Simonton1991829, Simonton1977791} used
time-series analysis to measure the creative productivity of composers
based on their music and popularity. Kozbelt~\cite{Kozbelt01012009,
  Kozbelt01012007} also analyzed productivity, but based on the
measure of performance time of the compositions and investigated the
relation between productivity and versatility. More recent
works~\cite{Kranenburg2004, Kranenburg2007} used machine-learning
algorithms to recognize musical styles of selected compositions.

In contrast to the studies above, we are not interested in applying
statistical analysis to music but on characterizing composers by
identification of scores based on their styles. On the other hand,
automatic information retrieval --- that has been applied to music ---
is not common in philosophy. The method proposed here is a way to
analyze both fields independently of the nature of their works ---
i.e.\ music pieces and textual documents --- but based on a
well-formed opinion of reviewers or critics on these fields and their
analysis of temporal evolution.

\section{Mathematical Description}

Each composer or philosopher and their characteristics
(i.e.\ opposition and skewness) are defined for each pair of
subsequent composers or philosophers along time. Therefore, the choice
of composers and philosophers is crucial for the time-evolution
analysis.  A sequence $S$ of $C$ music composers and $P$ philosophers
was chosen based on their relevance in each period of classical music
and western philosophy history, respectively. The set of $C$
measurements defined a $C$-dimensional space, henceforth referred to
as the \emph{musical space}. Likewise, we defined a $P$-dimensional
\emph{philosophical space} based on $P$ measurements.

The vector $\vec{v}_i$ for each composer or philosopher $i$ defines a
corresponding \emph{composer state} in the music space or
\emph{philosopher state} in the philosophy space. For the set of $C$
composers and $P$ philosophers, we defined the same relations
summarized in Table \ref{tab:tableRelations}. Some details about these
relations are worth noting from Figure \ref{fig.1} for composers, and
similar remarks apply to philosophers. Given a set of $C$ composers as
a time-sequence $S$, the \emph{average state} at time $i$ is defined
as $\vec{a}_i$. The \emph{opposite state} is defined as the
``counterpoint'' of a music state $\vec{v}_i$, considering its average
state: everything running along the opposite direction of $\vec{v}_i$
is understood as opposition. In other words, any displacement from
$\vec{v}_i$ along the direction $\vec{r}_i$ is a \emph{contrary move},
and any displacement from $\vec{v}_i$ along the direction $-\vec{r}_i$
is an \emph{emphasis move}. Given a composer state $\vec{v}_i$ and its
opposite state $\vec{r}_i$, the \emph{opposition vector} $\vec{D}_i$
is defined.

\begin{table}
  \caption{\label{tab:tableRelations}Description of mathematical
    relations for each composer or philosopher $i$, $j$ and $k$, given
    a set of $C$ composers or $P$ philosophers as a time-sequence $S$.
    The \emph{opposition index} $W_{i,j}$ quantifies how much a
    composer or philosopher state $i$ is opposed to $j$. The
    \emph{skewness index} $_{i,j}$ quantifies the extent to which the
    new composer or philosopher state $j$ departs from the
    corresponding opposition state. And the \emph{counter-dialectics}
    index quantifies how much a state $k$ could be considered a
    \emph{synthesis} between the \emph{thesis} state $i$ and the
    \emph{antithesis} state $j$.}

  \begin{ruledtabular}
    \begin{tabular}{ll}
      \\ 
      Average state      & $\vec{a}_i = \frac{1}{i}\sum_{k=1}^i\vec{v}_k.$ 
      \\ \\
      Opposite state     & $\vec{r}_i = \vec{v}_i + 2(\vec{a}_i - \vec{v}_i)$ 
      \\ \\
      Opposition vector  & $\vec{D}_i=\vec{r}_i - \vec{v}_i$ 
      \\ \\
      Composer or \\ 
      philosopher state \\move & $\vec{M}_{i,j} = \vec{v}_j - \vec{v}_i$ 
      \\ \\
      Opposition index  & $W_{i,j} = \frac{\left< \vec{M}_{i,j}, \vec{D}_i\right>}
      {||\vec{D}_i||^2}$ 
      \\ \\
      Skewness index     & $s_{i,j} = \sqrt{\frac{|\vec{v}_i-\vec{v}_j|^2
          |\vec{a}_i-\vec{v}_i|^2 - 
          [(\vec{v}_i-\vec{v}_j) . 
            (\vec{a}_i-\vec{v}_i)]^2}
        {|\vec{a}_i-\vec{v}_i|^2}}$ 
      \\ \\
      Counter-dialectics \\ 
      index              & $d_{i \rightarrow k} = 
      \frac{|\left< \vec{v}_j-\vec{v}_i,\vec{v}_k \right> + 
        \frac{1}{2}\left<\vec{v}_i-\vec{v}_j, \vec{v}_i+\vec{v}_j\right>|}
           {|\vec{v}_j-\vec{v}_i|}$
           \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.35\textwidth]{phil-space-desc_}
  \end{center}
  \caption{\it Graphical representation of the measures derived from a 
    \emph{composer state move}.}
  \label{fig.1}
\end{figure}

For the time-sequence $S$ relations between pairs of composers can be
defined, as follows. The move between two successive \emph{composer
  states} at time $i$ and $j$ corresponds to the $\vec{M}_{i,j}$
vector extending from $\vec{v}_i$ to $\vec{v}_j$. Given the composer
state we can quantify the intensity of opposition by the projection of
$\vec{M}_{i,j}$ along the opposition vector $\vec{D}_i$, normalized,
yielding the \emph{opposition index} $W_{i,j}$. With the same composer
state, the \emph{skewness index} $s_{i,j}$ is the distance between
$\vec{v}_j$ and the line $L_i$ defined by the vector $\vec{D}_i$, thus
quantifying the extent into which the new composer state departs from
the corresponding opposition state. A relationship between a triple of
successive composers can also be defined. Taking $i$, $j$ and $k$ as
the \emph{thesis}, \emph{antithesis} and \emph{synthesis} states, the
\emph{counter-dialectics index} $d_{i \rightarrow k}$ was defined by
the distance between the composer state $\vec{v}_k$ and the middle
line $ML_{i,j}$ defined by the thesis and antithesis, as shown in
Figure \ref{fig.2}. In higher dimensional musical or philosophical
spaces, the middle-hyperplane defined by the points which are at equal
distances to both $\vec{v}_i$ and $\vec{v}_j$ should be used instead
of the middle line $ML_{i,j}$. The proposed equation for
counter-dialectics scales to hyperplanes.  The counter-dialectics
index is used here, instead of a dialectics index, to maintain
compatibility with the use of a distance from point to line as adopted
for the definition of skewness.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.35\textwidth]{dialetica_.eps}
  \end{center}
  \caption{\it Graphical representation of the quantification of
    dialectics.}
  \label{fig.2}
\end{figure}

\section{Characteristics}

To create the music and philosophy spaces we derived eight variables
corresponding to distinct characteristics commonly found in music
compositions and works by philosophers. While the selected
characteristics cannot be considered a summary of all the relevant
features in music and philosophy history, they are initial indicators
that reveal differences in style of composers and philosophers. We
emphasize that the focus of this work is not on the specific
characteristics used or the scores assigned, which can be disputed,
but on the techniques for a quantitative analysis.

\subsection{Musical Characteristics}

These characteristics are related to basic elements of music ---
melody, harmony, rhythm, timbre, form and
tessitura~\cite{BennettHistory} --- in addition to non-musical issues
like historical events that influenced compositions, such as the
importance of the Church. The eight characteristics are listed
below:\\

{\bf \em{ Sacred - Secular}} ($S$-$S_c$): the sacred or religious
music is composed through religious influence or used for its
purposes. \textit{Masses}, \textit{motets} and hymns, dedicated to the
Christian liturgy, are well-known examples~\cite{Lovelock}. Secular
music has no or minimal relation with religion and includes popular
songs like Italian madrigals and German
\textit{lieds}~\cite{BennettHistory}.

{\bf \em{ Short duration - Long duration}} ($D_s$-$D_l$): compositions
quantified as short duration have a few minutes of execution. Long
duration compositions have at least 20 minutes of execution. The same
criterion was adopted by Kozbelt~\cite{Kozbelt01012009,
  Kozbelt01012007} in his analysis of time execution.

{\bf \em{ Harmony - Counterpoint}} ($H$-$C$): harmony regards the
vertical combination of notes, while counterpoint focuses on
horizontal combinations~\cite{BennettHistory}.

{\bf \em{ Vocal - Instrumental}} ($V$-$I$): compositions using just
vocals (e.g.\ \emph{cantata}) or exclusively instruments
(e.g.\ \emph{sonata}). Note the use of vocals over instruments on
Sacred compositions~\cite{Lovelock}.

{\bf \em{ Non-discursive - Discursive}} ($D_n$-$D$): compositions
based or not on verbal discourse, like programmatic music or Baroque
rhetoric, where the composer wants to ``tell a story'' invoking images
to the listeners mind~\cite{BennettHistory}. Its contrary part is
known as \textit{absolute music} where music was aimed to be
appreciated simply by what it is.

{\bf \em{ Motivic Stability - Motivic Variety}} ($M_s$-$M_v$): motivic
pieces present equilibrium between repetition, reuse and variation of
melodic motives. Bach is noticeable for his \textit{variation} of
motives, contrasting with the constantly inventive use of new
materials by Mozart~\cite{Webern}.

{\bf \em{ Rhythmic Simplicity - Rhythmic Complexity}} ($R_s$-$R_c$):
presence or not of polyrhythms, the use of independent rhythms at the
same time --- also known as \textit{rhythmic
  counterpoint}\cite{BennettHistory} --- a characteristic frequently
found in Romanticism and the works of 20th-century composers like
Stravinsky.

{\bf \em{ Harmonic Stability - Harmonic Variety}} ($H_s$-$H_v$): rate
of tonality change along a piece or its stability. After the highly
polyphonic development in Renaissance, Webern regarded Beethoven as
the composer who returned to the maximum exploration of harmonic
variety~\cite{Webern}.

\subsection{Characteristics in Philosophy}

We derived eight variables corresponding to some of the most recurrent
philosophical issues~\cite{Russel,Papineau,Deleuze2}. Each of these
variables, which defined an axis in the philosophy space, are briefly
described in the following. \\

{\bf \em{ Rationalism - Empiricism}} (\emph{R-E}): the rationalists
claim that the human acquaintance of knowledge/concepts is
significantly independent of sense experience. Empiricists understand
sensory experience as the main way to gain knowledge. Empiricism is in
the origin of the scientific method where knowledge must be based on
sensible observation of the world instead of faith or
intuition. Frequently, rationalists understand the world as affected
by intrinsic properties of the human brain, in contrast to the
empiricist approach where the world would imprint itself onto our
minds (the principle of \emph{``tabula rasa''}). For rationalists,
deduction is the superior method for investigation and privileges the
reason instead of experience as the source of true
knowledge. Historically, Descartes, Espinoza and Leibniz introduced
the rationalism in modern philosophy.

{\bf \em{ Essence - Existence}} (\emph{E-E}): An existence-based
understanding of the world has its basis on the fact that things are
as an existent unit. In Existentialism, the life of an individual is
determined by its self, that constitutes its essence, and not a
predefined essence that defines what is to be human. Essence focuses
on a substance (e.g.\ intellectual) that precedes existence
itself. For essentialists, any specific entity has essential
characteristics necessary for its function and identity. For example,
to have two legs and the ability to run is not an essential
characteristic defining humans, because other animals have the same
characteristics. Plato was one of the first essentialists while
Nietzsche and Kierkegaard were fundamental to Existentialism in the
19th century.

{\bf \em{ Monism - Dualism}} (\emph{M-D}): Dualism requires the
division of the human person into two or more domains, such as matter
and soul. Monism is based on a unique ``category of being''. Plato is
a recognized dualist while his disciple, Aristotle, was a notable
monist.

{\bf \em{ Theocentrism - Anthropocentrism}} (\emph{T-A}): In
theocentrism, God is the most important thing in the universe. The
anthropocentric view has man as prevalent, with Nietzsche being its
main representative. It is important to distinguish theocentrism from
deism, where the figure of God is considered as an abstract entity, as
for Espinoza, but does not play a central role in the universe.

{\bf \em{ Holism - Reductionism}} (\emph{H-R}): Reductionism attempts
to explain the world in terms of simple components and their emerging
properties. Holists focus on the fact that the whole is more than its
constitutive parts.

{\bf \em{ Deductionism - Phenomenology}} (\emph{D-P}): Phenomenology
relies on systematic reflection of consciousness and what happens in
conscious acts. Deductionism is based on deriving conclusions from
axiomatic systems.

{\bf \em{ Determinism - Free Will}} (\emph{D-F}): Free will assumes
that humans make choices, which are not predetermined. Determinism
understands that every event is fatidic, e.g.\, perfectly determined
by prior states.

{\bf \em{ Naturalism - Mechanism}} (\emph{N-M}): Methodological
naturalism is the thinking basis of modern science, i.e. hypotheses
must be argued and tested in terms of natural laws. Mechanism attempts
to build explanation using logic-mathematical processes.

\subsection{Bootstrap method for sampling}
\label{bootstrap}

To eliminate the bias intrinsic in a small sample group, we used a
bootstrap method for generating \emph{artificial composers and
  philosophers} contemporaries of those seven chosen. The bootstrap
routine generated new scores $\vec{r}$, which are not totally random
because they follow a probability distribution that models the
original $n = 7$ scores, given by $p(\vec{r}) = \sum^n_{i=1}
e^{\frac{d_i}{2\sigma^2}}$ where $d_i$ is the distance between a
random score $\vec{r}$ and the original score chart. For each step a
value $p(\vec{r})$ is generated and compared with a random normalized
value, as in the Monte Carlo~\cite{Robert2011} method to choose a set
of samples. These samples simulate new randomized composers and
philosophers score charts --- while preserving the historical
influence of the main 7 original names in each field. Higher values of
$p(\vec{r})$ imply a stronger influence of the original scores over
$\vec{r}$. For the analysis we used 1000 bootstrap samples obtained by
the bootstrap process together with the original scores, taking
$\sigma = 1.1$. Other values for $\sigma$ were used yielding
distributions with bootstrap samples that did not affect the music or
philosophy space substantially.

\section{Results and Discussion}

Memorable composers were chosen as key representatives of classical
music, which we believe had impact on their contemporaries, thus
creating a concise parallel with music history. The chronological
sequence is presented in Table \ref{tab:table0} with each composer
related to his historical period. The same was done for philosophy
where a set of seven philosophers were chosen spanning the period from
Classical Greece until contemporary times, and ordered chronologically
as: Plato, Aristotle, Descartes, Espinoza, Kant, Nietzsche and
Deleuze, as shown in Table \ref{tab:table0b}.

 \begin{table}[ht]
   \caption{\label{tab:table0} Sequence of music composers ordered
     chronologically with the period each represents.}
   \begin{tabular}{|l||l|}
     \hline
     Composer        & Movement \\ \hline
     Monteverdi      & Renaissance \\
     Bach            & Baroque \\
     Mozart          & Classical \\
     Beethoven       & Classical $\to$ Romantic \\
     Brahms          & Romantic \\
     Stravinsky      & 20th-century \\
     Stockhausen     & Contemporary\\
     \hline
   \end{tabular}
 \end{table}

\begin{table}[ht]
  \caption{\label{tab:table0b} Sequence of philosophers ordered
    chronologically with the period each represents.}
  \begin{tabular}{|l||l|}
    \hline
    Philosopher     & Era \\ \hline
    Plato           & Ancient \\
    Aristotle       & Ancient  \\
    Descartes       & 17th-century  \\
    Espinoza        & 17th-century  \\
    Kant            & 18th-century  \\
    Nietzsche       & 19th century  \\
    Deleuze         & 20th-century  \\
    \hline
  \end{tabular}
\end{table}

The quantification of the eight characteristics for music and
philosophy was performed jointly by three of the authors of this
article, based on research of history of music and western
philosophy. The scores shown in Tables \ref{tab:tableAmus} and
\ref{tab:tableAphi} for philosophers and composers, respectively, were
numerical values between 1 and 9. Values closer to 1 reveal the
composer or philosopher tended to the first element of each
characteristic pair, and vice-versa.

\begin{table}[ht]
  \caption{\label{tab:tableAmus} Quantification of eight music
    characteristics for each of the seven composers.}
  \begin{ruledtabular}
    \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
      \footnotesize Composers    & \tiny  $S$-$S_c$ 
      & \tiny  $D_s$-$D_l$ & \tiny  $H$-$C$ & \tiny  $V$-$I$ 
      & \tiny  $D_n$-$D$ & \tiny  $M_s$-$M_v$ & \tiny  $R_s$-$R_c$
      & \tiny  $H_s$-$H_v$  \\
      \hline
   \footnotesize Monteverdi  & 3.0 & 8.0 & 5.0 & 3.0 & 7.0 & 5.0 & 3.0 & 7.0 
   \\
   \footnotesize Bach        & 2.0 & 6.0 & 9.0 & 2.0 & 8.0 & 2.0 & 1.0 & 5.0 
   \\
   \footnotesize Mozart      & 6.0 & 4.0 & 1.0 & 6.0 & 6.0 & 7.0 & 2.0 & 2.0 
   \\
   \footnotesize Beethoven   & 7.0 & 8.0 & 2.5 & 8.0 & 5.0 & 4.0 & 4.0 & 7.0 
   \\
   % Chopin & 9.0 & 3.0 & 3.0 & 9.0 & 5.5 & 8.0 & 7.0 & 8.0 \\
   \footnotesize Brahms      & 6.0 & 6.0 & 4.0 & 7.0 & 4.5 & 6.5 & 5.0 & 7.0 
   \\
   \footnotesize Stravinsky  & 8.0 & 7.0 & 6.0 & 7.0 & 8.0 & 5.0 & 8.0 & 5.0 
   \\
   \footnotesize Stockhausen & 7.0 & 4.0 & 8.0 & 7.0 & 5.0 & 8.0 & 9.0 & 6.0 
   \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

\begin{table}%\footnotesize%\scriptsize%\tiny
  \caption{\label{tab:tableAphi} Quantification of eight philosophy
    characteristics for each of the seven philosophers.}

  \begin{ruledtabular}
    \begin{tabular}{|l||c|c|c|c|c|c|c|c|}
      Philosophers & R-E & E-E & M-D & T-A & H-R & D-P & D-F & N-M \\ \hline
      Plato        & 3.0 & 3.5 & 9.0 & 5.0 & 4.5 & 3.5 & 5.0 & 4.5 \\
      Aristotle    & 8.0 & 7.5 & 7.0 & 5.5 & 7.5 & 8.0 & 2.5 & 2.5 \\
      Descartes    & 1.5 & 2.5 & 9.0 & 6.5 & 7.0 & 2.5 & 7.5 & 7.5 \\
      Espinoza     & 8.0 & 2.0 & 1.0 & 5.0 & 2.0 & 3.0 & 1.0 & 1.0 \\
      Kant         & 7.0 & 2.5 & 8.5 & 6.5 & 4.5 & 3.5 & 7.5 & 5.0 \\
      Nietzsche    & 7.5 & 9.0 & 1.0 & 9.0 & 5.0 & 8.0 & 1.0 & 1.5 \\
      Deleuze      & 5.5 & 7.5 & 1.0 & 8.0 & 2.5 & 5.5 & 5.0 & 6.0 \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

This data set defines an 8-dimensional space for music or philosophy
where each dimension corresponds to a characteristic that applies to
all 7 composers or philosophers. Such small data sets are not adequate
for statistical analysis, which could be biased. This is the reason
why we used the bootstrap method for sampling described in
section~\ref{bootstrap}.

For the extended data set after applying the bootstrap method, Pearson
correlation coefficients between the eight characteristics chosen are
given in Table \ref{tab:tableBmus} for composers and in Table
\ref{tab:tableBphi} for philosophers. The coefficient was 0.69 for the
pairs $S$-$S_c$ (Sacred or Secular) and $V$-$I$ (Vocal or
Instrumental), which indicates that sacred music tends to be more
vocal than instrumental. The coefficient 0.56 for the pairs $S$-$S_c$
and $R_s$-$R_c$ (Rhythmic Simplicity or Complexity) also shows that
this genre does not commonly use polyrhythms. A negative coefficient
of -0.33 for the pair $V$-$I$ and $D_n$-$D$ (Non-discursive or
Discursive) indicated that composers who used just voices on their
compositions also preferred programmatic music techniques such as
baroque rhetoric. Strong correlations were also observed for
philosophers. For instance, the Pearson coefficient of $-0.46$ for
\emph{R-E} and \emph{N-M} suggests that rationalists tend to be also
mechanists. An even stronger correlation of $0.74$, now positive, is
observed between \emph{E-E} and \emph{D-P}, i.e. existentialists also
tend to be phenomenologists, as could be expected. Other important
correlations appeared between \emph{D-F} and \emph{N-M} (coefficient =
$0.61$) and between \emph{M-D} and \emph{D-F}, which seems to be
directly implied by religious background.

\begin{table}[ht]
  \caption{\label{tab:tableBmus}Pearson correlation coefficients
    between the eight musical characteristics. The entries with
    absolute values $\geq$ 0.30 have been highlighted.}
  \begin{ruledtabular}
    \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
      -    & \tiny  $S$-$S_c$ & \tiny  $D_s$-$D_l$ & \tiny  $H$-$C$ 
      & \tiny  $V$-$I$ & \tiny  $D_n$-$D$ & \tiny  $M_s$-$M_v$ 
      & \tiny  $R_s$-$R_c$ & \tiny  $H_s$-$H_v$  \\ \hline

      \tiny$S$-$S_c$ 
      & - & -0.2 & -0.06 & \bf{0.69} & -0.18 & 0.19  & \bf{0.56} & -0.16 
      \\
      \tiny$D_s$-$D_l$ 
      & - & - & -0.14 & -0.13 & 0.2 & \bf{-0.48} & -0.2 & \bf{0.37}
      \\
      \tiny$H$-$C$ 
      & - & - & - & -0.23 & 0.26 & 0.05 & \bf{0.46} & 0.03 
      \\
      \tiny$V$-$I$ 
      & - & - & - & - & \bf{-0.33} & 0.17 & \bf{0.42} & -0.06 
      \\
      \tiny$D_n$-$D$ 
      & - & - & - & - & - & -0.3 & 0.02 & -0.22 
      \\
      \tiny$M_s$-$M_v$ 
      & - & - & - & - & - & - & 0.26 & -0.15
      \\
      \tiny$R_s$-$R_c$ 
      & - & - & - & - & - & - & - & -0.02 
      \\
      \tiny$H_s$-$H_v$ 
      & - & - & - & - & - & - & - & - 
      \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

\begin{table}\footnotesize
  \caption{\label{tab:tableBphi}Pearson correlation coefficients
    between the eight philosophical characteristics.  The entries with
    absolute values $\geq$ 0.35 have been highlighted.}
  \begin{ruledtabular}
    \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
      - & R-E & E-E & M-D & T-A & H-R & D-P & D-F & N-M \\ \hline
      R-E & - & \bf{0.37} & -0.23 & 0.15 & 0.1 & \bf{0.46} & -0.27 & \bf{-0.46} 
      \\
      E-E & - & - & \bf{-0.53} & 0.19 & 0.15 & \bf{0.74} & \bf{-0.61} & -0.3 
      \\
      M-D & - & - & - & \bf{-0.43} & \bf{0.41} & -0.3 & \bf{0.35} & 0.01 
      \\
      T-A & - & - & - & - & -0.21 & 0.06  & 0.19 & 0.26 
      \\
      H-R & - & - & - & - & - & 0.32 & -0.22 & -0.25 
      \\ 
      D-P & - & - & - & - & - & - & \bf{-0.63} & \bf{-0.47} 
      \\
      D-F & - & - & - & - & - & - & - & \bf{0.61}
      \\
      N-M & - & - & - & - & - & - & - & - 
      \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

>>>>

\section{Characterization of the discrete-time musical note}
In diverse artistic and theoretical contexts, music is though of as being constituted by units called notes and this units taken as "atoms" that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
Nowadays, this notes are undestood as a central element of certain musical paradigms. In a cognitive sense, the notes are seen as discretizations that eases and enrich the flow of information through music~\cite{Roederer, Lacerda}
Canonically, a musical note posses at least duration, colume, pitch and tibre~\cite{Lacerda}. These are qualities which can be maneged quantitatively dictated by the evenly time spaced sound samples.

@@

\subsection{Duration}
The sample frequency $f_s$ is defined as the number os samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds. A Musical note od duration $\Delta$ seconds is presented as a sequence of $\lfloor \Delta . f_s \rfloor $ samples. That is, taken the integer part of the multiplication, it is admited an error of at most $-\lambda_a$ seconds, which, for musical porpuses, are usually fine, for $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds. It is reasonable to state:

@@

\subsection{Volume}\label{subsec:volume}
The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics worked on section~\ref{sec:varInternas}. One can get variations of volume through the potency of the wave~\cite{Chowning}:

@@

The final volume is dependent on the speakers amplification of the signal. Thus, what matters is the relative potency of a note in relation to the ones around it or the potency of a music section in relation to the rest. Differences in volume are measured in decibels, and these are calculated directly form the amplitudes through energy or potency:

\begin{equation}\label{decibels}
V_{dB}=10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)}
\end{equation}

The quantity $V_{dB}$ has the decibel unit ($dB$). 
To each $10dB$ it is associated a "doubled volume".
Handy references are the $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}. Other useful references are que equivalent in $dB$ of doubling amplitude or potency:

\begin{equation}\label{eq:ampVol}
\begin{split}
t_i^{'}=2 . t_i \Rightarrow pot(T^{'}_i)=4 . pot(T_i) \Rightarrow \\ \Rightarrow V^{'}_{dB}=10log_{10} 4 \approx 6 dB
\end{split}
\end{equation}

\begin{equation}\label{eq:potVol}
\begin{split}
pot(T^{'}_i)=2 pot(T_i) \Rightarrow \\ \Rightarrow V^{'}_{dB}=10log_{10} 2 \approx 3 dB
\end{split}
\end{equation}

and the amplitude gain for a sequence whose volume has been doubled ($10dB$):

\begin{equation}\label{eq:dobraVol}
\begin{split}
10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \\ \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_a \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \\
\therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3,16t_i
\end{split}
\end{equation}

Thus, it is necessary a little bit more than to triplicate the amplitude for a doubled volume. This values are guides for the increases and decreases on the absolute values on the sample sequences with musical purposes. The direct conversion from decibels to amplitude gain (or attenuation) is:

\begin{equation}\label{ampDec}
A = 10^{\frac{V_{dB}}{20}}
\end{equation}

Where $A$ is the mutiplicative factor that relates the amplitudes before and after the amplification.

\subsection{Pitch}
Recapitulating, the musical particle (note) is a sequence $T_i$ which duration and volume corresponds to the size of sequence and the amplitude of its samples. The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$. This duration, multiplied by the sampling frequency $f_s$ results on the number of samples of the cycle $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$.

For didatic reasons, be $f_0$ such that it divides $f_s$ and $\lambda_{f_0}$ results an integer. If $T_i^f$ is a sonic sequence of fundamental frequency $f$, then:

\begin{equation}\label{periodicidade}
     T^f_i=\left\{ t_i^f \right\}=\left\{ t^f_{i+\lambda_{f}}  \right\}= \left\{ t^f_{i+\frac{f_a}{f}} \right\}
\end{equation}

In the next secion, frequencies $f$ that does not divide $f_s$ will be considered. This restriction does not imply in lost of generality of this section's content.

\subsection{Timbre}
While the period of the wave corresponds to a fundamental frequency, the trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{The timbre is a subjective and complex characteristic. Physicaly, the timbre is multidimentionsl and given by the temporal dinamics behavior of enegy in spectral components that are harmonic or noisy. Beyond that, the word timbre is used to designate different things: one same note has different timbres, a same instrument has different timbres, two intruments of the same family possesses the same timbre that blends them in the same family and different timbres as they are different instruments. It is worth to mention that not all that is associated to timbre is manifested in spectral traces, even cultural or circunstancial aspects alter our perception of timbre}. Musically, it matters that sonic spectra with minimum differences result in timbres with crucial expressive differences and that, hence, diffetent timbres can be produced by using different spectra\cite{Roederer}.

The simplest case (and most important, as the following texts shows) in that of the spectrum that consists only of the fundamental frequency $f$ itself. This is the sinusoid, frequency in pure oscilatory movement called 'simple harmonic movement'. Be $S_i^f$ a sequence whose samples $s_i^f$ describes a sibusoid of frequency $f$:

\begin{equation}\label{senoide}
     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(2\pi \frac{i}{\lambda_f} \bigr)  \Bigr\} = \Bigl\{ \sin\bigl(2\pi f \frac{i}{f_a}\bigr)  \Bigr\} 
\end{equation}
